---
title: "Breast Cancer R Analysis Project"
author: Blair Steele
course: BINF 5003 - Data Mining
student ID: n01606240
date: April 16, 2026
---

```{r}
# save.image("my_workspace.RData")
# save.image("BcancerRAnalysisProject_workspace.RData")
# load("BcancerRAnalysisProject_workspace.RData")

```

```{r}
# install.packages("datasauRus")
# install.packages(c("ggpubr", "ggsignif", "ggrepel"))
# install.packages("pheatmap")

library(tidyverse)
library(dplyr)
library(survminer)
library(survival)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)
library(limma)
library(tibble)
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
library(datasauRus)

```

----------------------------
# CLINICAL DATA PRE-PROCESSING, TRANSFORMATION, AND INTEGRATION
----------------------------

# Set Working Dictionary
setwd("/Users/BGS/Desktop/🧬 CLINICAL BIOINFORMATICS - HUMBER (LU = Dec.4.2025)/★ SCHOOL/★ SEMESTER 1/BINF 5003 - Data Mining, Modeling and Biostatistics/ASSIGNMENTS ☆ Data Mining, Modeling and Biostatistics - BINF 5003/RNA Analysis Assignment - Breast Cancer 2")

# Check if the file exists in folder
file.exists("/Users/BGS/Desktop/🧬 CLINICAL BIOINFORMATICS - HUMBER (LU = Dec.4.2025)/★ SCHOOL/★ SEMESTER 1/BINF 5003 - Data Mining, Modeling and Biostatistics/ASSIGNMENTS ☆ Data Mining, Modeling and Biostatistics - BINF 5003/RNA Analysis Assignment - Breast Cancer 2") #TRUE file exists here
read.table("Human__TCGA_BRCA__MS__Clinical__Clinical__01_28_2016__BI__Clinical__Firehose.tsi")

# Load clinical data as dataframe/table
matrix_table <- read.table(
  "Human__TCGA_BRCA__MS__Clinical__Clinical__01_28_2016__BI__Clinical__Firehose.tsi",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE

)

exists("matrix_table") #TRUE
list.files()

dim(matrix_table) #20rows (clinical attributes) 1,098columns (# of patients)
head(matrix_table)
head(matrix_table[, 1:5]) #First 6 rows & 5 columns, by default
colnames(matrix_table)
rownames(matrix_table)

col2 <- matrix_table[, 1]
head(col2)  

View(matrix_table)
dim(matrix_table) #20 x 1,098 #NOT TIDY FORMAT

# Save Matrix
write.csv(matrix_table, "matrix_table.csv", row.names = FALSE)

# Load Matrix
matrix_table <- read.csv("matrix_table.csv")

----------------------------
# DATA WRANGLING FOR MATRIX TABLE

# Transpose matrix_table
rownames(matrix_table) <- matrix_table [,1] #this makes the attributes into the rows
new_matrix <- matrix_table [, -1] #first column (i.e., attribute names) removed
transposed_matrix <- as.data.frame(t(new_matrix)) #transpose the matrix
transposed_matrix$Patient_Barcode <- rownames(transposed_matrix) #add new column = "patient barcode"
rownames(transposed_matrix) <- NULL

# Make patient barcode the second column (i.e, after row number)
cols <- colnames(transposed_matrix)
cols <- cols[cols != "Patient_Barcode"] #removes the column from the position it is now in
cols <- append(cols, "Patient_Barcode", after = 0) #puts column in index 1
transposed_matrix <- transposed_matrix[, cols]

# Change all column names to lowercase.
colnames(transposed_matrix) <- tolower(colnames(transposed_matrix))

# Standardize the barcode (remove . and replace with -)
transposed_matrix$patient_barcode <-
      toupper(gsub("\\.", "-", trimws(transposed_matrix$patient_barcode)))

# Check for missing values/NA's in patient barcode column in transposed matrix table.
sum(is.na(transposed_matrix$patient_barcode)) #returns 0 N/A values in patient barcode column in transposed matrix table.

# Check for spaces at the end of variables
grep("^\\s|\\s$", colnames(transposed_matrix), value = TRUE) #returns character(0) = means no column names have any leading or trailing spaces.

# Fix row numbering after wrangling
rownames(transposed_matrix) <- NULL

# Sort transposed_matrix table by patient barcode
# transposed_matrix <- transposed_matrix[
#       order(transposed_matrix$patient_barcode),
# ]

# Summary of categorical variables in transposed_matrix
categorical_cols <- sapply(transposed_matrix, is.character) #identifies factor/character columns
print(categorical_cols)
lapply(transposed_matrix[, categorical_cols], unique) #identifies unique values in each categorial column
lapply(transposed_matrix[, categorical_cols], table) #identifies frequency of each unique value in each categorical column
# Notice "NAs"

# CHECK FOR TYPOS/INCONSISTENT CODING IN CATEGORICAL DATA (race, gender, pathogical stage)
# Inconsistent coding may include entries such as "Male", "male", "M", "Female", "F", "Femlae" etc.

# Check for missingness (i.e., missing or N/A values)
sum(is.na(transposed_matrix)) #4144 NAs 

# Change format from TCGA.5L.AAT0 to TCGA-5L-AAT0
transposed_matrix$patient_barcode <- toupper(gsub("\\.", "-", trimws(transposed_matrix$patient_barcode))) #now the patient barcodes are in the same format in both tables

dim(transposed_matrix) #1097 x 21
View(transposed_matrix) #new 'transposed clinical matrix' table after wrangling

# Save Matrix
write.csv(transposed_matrix, "transposed_matrix.csv", row.names = FALSE)

# Load Matrix
transposed_matrix <- read.csv("transposed_matrix.csv")

----------------------------
# COMBINE ALL TSV FILES 

# Set WD
setwd("/Users/BGS/Desktop/🧬 CLINICAL BIOINFORMATICS - HUMBER (LU = Dec.4.2025)/★ SCHOOL/★ SEMESTER 1/BINF 5003 - Data Mining, Modeling and Biostatistics/ASSIGNMENTS ☆ Data Mining, Modeling and Biostatistics - BINF 5003/RNA Analysis Assignment - Breast Cancer 2/clinical.project-tcga-brca.2026-02-12")

# List all .tsv files in the folder
files <- list.files(pattern = "\\.tsv$")
print(files)
View(files)

# READ .tsv FILES
all_tsv_files <- lapply(files, function(f) {
      all_tsv_df <- read_tsv(f, show_col_types = FALSE)
      colnames(all_tsv_df) <- trimws(colnames(all_tsv_df)) #trimws removes white spaces

      if ("cases.submitter_id" %in% colnames(all_tsv_df)) {
        all_tsv_df <- all_tsv_df %>%
            filter(!is.na(cases.submitter_id) & cases.submitter_id != "") %>% #Removes missing NA values. "cases.submitter_id" is essentially the patient barcode (rename later)
            group_by(cases.submitter_id) %>%
            slice(1) %>% #ensures one row per patient
            ungroup()
      }
      all_tsv_df 

})

# Merge all .tsv files by "cases.submitter_id"
combined_patient_tsv_files <- Reduce(function(x, y) {
      full_join(x, y, by = "cases.submitter_id")
}, all_tsv_files)

# Ensure duplicated columns, if any, are removed
combined_patient_tsv_files <-
      combined_patient_tsv_files[, !duplicated(colnames(combined_patient_tsv_files))]

# Rename "cases.submitter_id" to "patient_barcode"
colnames(combined_patient_tsv_files)[colnames(combined_patient_tsv_files) == "cases.submitter_id"] <- "patient_barcode"

# Check if any NA's are in the patient barcode column
any(is.na(combined_patient_tsv_files$patient_barcode)) #FALSE - no NA's

# Move "patient_barcode" to index 1
cols <- colnames(combined_patient_tsv_files)
cols <- cols[cols != "patient_barcode"]
cols <- c("patient_barcode", cols)
combined_patient_tsv_files <- combined_patient_tsv_files[, cols]

# Check for spaces in column names 
grep("^\\s|\\s$", colnames(combined_patient_tsv_files), value = TRUE) #returns character(0) = means no column names have any leading or trailing spaces.

# Note = Column names are already lowercase (common practice)

head(combined_patient_tsv_files)
View(combined_patient_tsv_files)
colnames(combined_patient_tsv_files)
dim(combined_patient_tsv_files) #1098 x 543

# NOTE = the combined_patient_tsv_files is now in tidy format (row = patient, column = variables), and does not need to be transposed. 

# Save Matrix
write.csv(combined_patient_tsv_files, "combined_patient_tsv_files.csv", row.names = FALSE)

# Load Matrix
combined_patient_tsv_files <- read.csv("combined_patient_tsv_files.csv")

----------------------------
# Number of Males and Females in the cohort 
table(combined_patient_tsv_files$demographic.gender) #Males = 12, Females = 1085

# Check for inconsistent coding entries such as "Male", "male", "M", "Female", "F", "Femlae" etc.
unique(combined_patient_tsv_files$demographic.gender) 
# Only "female", "male", and "'--"  exit

----------------------------
# MERGE BOTH TRANSPOSED MATRIX (CLINICAL DATA) & COMBINED PATIENT TSV FILE MATRIX

clinical_cols <- setdiff(colnames(transposed_matrix), "patient_barcode")  #identifies all columns of the transposed matrix except the "patient_barcode" column

new_clinical_cols <- clinical_cols[!clinical_cols %in% colnames(combined_patient_tsv_files)] #ensures that only unique clinical data from the transposed matrix is added to the combined_patient_tsv_files matrix

transposed_subset <- transposed_matrix %>%
      select(patient_barcode, all_of(new_clinical_cols)) #create a tidy subset of all unique variables + the patient barcodes

final_aggregated_patient_data <- combined_patient_tsv_files %>%
      left_join(transposed_subset, by = "patient_barcode") #matches variables to patient barodes

all_cols <- c("patient_barcode", new_clinical_cols,
      setdiff(colnames(final_aggregated_patient_data), c("patient_barcode", new_clinical_cols)))

final_aggregated_patient_data <- final_aggregated_patient_data[, all_cols] #reorders the columns

dim(final_aggregated_patient_data) #1098 x 536
View(final_aggregated_patient_data)

# Save Matrix
write.csv(final_aggregated_patient_data, "final_aggregated_patient_data.csv", row.names = FALSE)

# Load Matrix
final_aggregated_patient_data <- read.csv("final_aggregated_patient_data.csv")

----------------------------
# EXPLORATORY ANALYSIS - VISUALIZATION OF THE RELATIONSHIPS BETWEEN IMPORTANT CLINICAL VARIABLES

# AGE DISTRIBUTION

# Convert to numeric
age <- as.numeric(transposed_matrix[ ,"years_to_birth"])

# Remove N/A's 
age <-age[!is.na(age)]

# Plot Age Distribution Histogram
hist(age,
     main = "Distribution of Age",
     xlab = "Age (In Years)",
     col = "yellowgreen",
     breaks = 20)

# Save Object
png("age_histogram.png", width = 800, height = 800)
plot(x, y)
dev.off()

# Youngest and Oldest Patients
age <- as.numeric(transposed_matrix$years_to_birth)
age <- age[!is.na(age)] #remove NAs 
min(age) #26
max(age) #90
mean(age) #58.59
median(age) #59
sd(age) #13.19

----------------------------
# TUMOR PURITY DISTRIBUTION

# Convert to numeric
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"])

# Remove N/A's 
purity <-purity[!is.na(purity)]

# Plot Tumor Purity Histogram
hist(purity, 
    main = "Tumor Purity Distribution", 
    xlab = "Purity", 
    col="#e69e41", 
    breaks =20)

----------------------------
# TUMOR PURITY vs. AGE

# Convert to numeric
age <- as.numeric(transposed_matrix[ ,"years_to_birth"])
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"])

purity_age_df <- data.frame(Age = age, TumorPurity = purity)
purity_age_df <- purity_age_df[complete.cases(purity_age_df), ] #this removes the rows that have an NA

# Plot Tumor Purity v. Age Plot
plot(purity_age_df$Age, purity_age_df$TumorPurity,
      xlab = "Age (Years)",
      ylab = "Tumor Purity",
      main = "Tumor Purity v. Age",
      pch =16, #plot point = solid circle
      col = "red") 

----------------------------
# TUMOR PURITY vs. OVERALL SURVIVAL

# Convert to numeric
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"]) #extracts tumor_purity column
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])

# Create dataframe
df_surv <- data.frame(Purity = purity, OverallSurvival = overall_survival)

# Remove N/A's
df_surv <- df_surv[!is.na(df_surv$Purity) & !is.na(df_surv$OverallSurvival), ]

# Plot Tumor Purity v. Overall Survival Plot
plot(df_surv$Purity, df_surv$OverallSurvival,
    xlab = "Tumor Purity",
    ylab = "Overall Survival (days)",
    main = "Tumor Purity v. Overall Survival",
    pch = 16, col = "#a9db2b")

# Add Linear Trend Line
abline(lm(OverallSurvival ~ Purity, data = df_surv), col= "#ff5500", lwd=2)

----------------------------
# TUMOR PURITY v. STAGE

stage <- unlist(transposed_matrix[ ,"pathologic_stage"]) #extracts pathological_stage column
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"]) #convert to numeric

# Create Dataframe
df_purity_stage <- data.frame(Stage=stage, Purity=purity)

# Remove N/A's
df_purity_stage <- df_purity_stage[!is.na(df_purity_stage$Stage) & !is.na(df_purity_stage$Purity), ]

# Plot Tumor Purity v. Stage Boxplot
boxplot(Purity ~ Stage, data=df_purity_stage, 
      main="Tumor Purity v. Stage", 
      col="powderblue")

----------------------------
# EXPLORATION OF PAM50 SUBTYPES

table(transposed_matrix$pam50) 
# Basal (147)  Her2 (67)  LumA(427)  LumB(186)
# Most patients have the Luminal A molecular subtype of breast cancer.

----------------------------
# SURVIVAL DISTRIBUTION

# Remove N/As
overall_survival <- overall_survival[!is.na(overall_survival)]

# Plot Survival Distribution Histogram
hist(overall_survival, 
      main = "Overall Survival Distribution", 
      xlab="Survival (days)", 
      col= "tomato2", 
      breaks=20)

----------------------------
# AGE vs. SURVIVAL OUTCOME

# Convert to numeric
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])
age <- as.numeric(transposed_matrix[ ,"years_to_birth"])

# Create Dataframe
df <- data.frame(Age = age, OverallSurvival = overall_survival)

# Remove N/As
df <- df[complete.cases(df), ]

# Plot Age v. Survival Outcome Plot
plot(df$Age, df$OverallSurvival,
    xlab = "Age",
    ylab = "Overall Survival (Days)",
    main = "Patient Age v. Survival Outcome",
    col = "#dc9107",
    pch = 16) 

# Add Trendline
abline(lm(OverallSurvival ~ Age, data=df), col="blue", lwd=2)

----------------------------
# OVERALL SURVIVAL BY STAGE

stage <- unlist(transposed_matrix[ ,"pathologic_stage"]) #extracts pathological_stage column
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])

# Create Dataframe 
df2 <- data.frame(Stage = stage, Survival = overall_survival)

# Plan Overall Survival by Stage Plot 
boxplot(Survival ~ Stage, data = df2,
        main = "Overall Survival by Stage",
        col = "mistyrose")

----------------------------
# NUMBER OF LYMPH NODES v. OVERALL SURVIVAL

lymph_nodes <- as.numeric(transposed_matrix[ , "number_of_lymph_nodes"])
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])

# Create dataframe
df_ln <- data.frame(LymphNodes = lymph_nodes, OverallSurvival = overall_survival)

# Remove N/As
df_ln <- df_ln[complete.cases(df_ln), ]

# Plot Number of Lymph Nodes v. Overall Survival Plot 
plot(df_ln$LymphNodes, df_ln$OverallSurvival,
    xlab = "Number of Lymph Nodes",
    ylab = "Overall Survival (Days)",
    main = "Number of Lymph Nodes v. Overall Survival",
    col = "#29aaef",
    pch = 16)

----------------------------
# PATHOLOGIC STAGE DISTRIBUTION

stage <- unlist(transposed_matrix[ ,"pathologic_stage"]) #extracts pathological_stage column
barplot(table(stage), main="Pathologic Stage Distribution", col="#039a9a", las=2)

----------------------------
# CORRELATION CHECK BETWEEN CONTINUOUS VARIABLES:

# Age v. Tumor Purity 
cor(age, purity, use="complete.obs") #0.082

# Age v. Overall Survival

# Tumor Purity v. Overall Survival

----------------------------
# PRE-PROCESSING & QUALITY ASSESSMENT OF RNA-SEQ GENE EXPRESSION DATA.

# getwd()

# LOAD RNA SEQUENCE DATA (.cct file)
lines <- readLines ("Human__TCGA_BRCA__UNC__RNAseq__GA_RNA__01_28_2016__BI__Gene__Firehose_RSEM_log2.cct", n = 5)
print(lines)

RNA_seq_data <- read.table("Human__TCGA_BRCA__UNC__RNAseq__GA_RNA__01_28_2016__BI__Gene__Firehose_RSEM_log2.cct", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Change "gene name" to "gene_name"
RNA_seq_data <- RNA_seq_data %>%
      rename('gene_name' = 'attrib_name')

# Fix format of patient barcodes from TCGA.A1.A0SD to TCGA-A1-A0SD (consistent across all datasets)
colnames(RNA_seq_data)[-1] <- gsub("\\.", "-", colnames(RNA_seq_data)[-1])

colnames(RNA_seq_data)
str(RNA_seq_data)
head(RNA_seq_data)
nrow(RNA_seq_data) #6,149 rows = 6,149 genes
View(RNA_seq_data)
summary(RNA_seq_data)

# RNA_seq_table is just RNA_seq_data without the gene_name column

# Convert all columns except attrib_name to NUMERIC
RNA_seq_table <- RNA_seq_data[, -1] #removes gene name column (i.e., the column with the names of the genes). 
RNA_seq_table <- data.frame(lapply(RNA_seq_table, function(x) as.numeric(as.character(x)))) #converts to numeric
RNA_seq_table <- as.matrix(RNA_seq_table) #converts to matrix

# Convert to matrix
RNA_seq_table <- as.matrix(RNA_seq_table)
class(RNA_seq_table) #check that it has been converted to a dataframe

# Standardize column names (replace . for - in patient barcodes)
colnames(RNA_seq_table) <- toupper(
      gsub("\\.", "-", trimws(colnames(RNA_seq_table)))
)

View(RNA_seq_table)

# Save Matrix as .csv
write.csv(RNA_seq_data, "RNA_seq_data.csv", row.names = FALSE)

# Load RNA_seq_data Matrix
RNA_seq_data <- read.csv("RNA_seq_data.csv")

write.csv(RNA_seq_table, "RNA_seq_table.csv", row.names = FALSE)

# Load RNA_seq_table Matrix
RNA_seq_table <- read.csv("RNA_seq_table.csv")

----------------------------

# CONVERT TIDY FORMAT: #3,233,933 rows?
# each row should have one observation (i.e., one gene per row)
# each column should have one variable

RNA_seq_tidy <- RNA_seq_data %>%
      pivot_longer(
      cols = -gene_name, #preserves the gene name
      names_to = "sample",
      values_to = "expression"
      ) %>%
      mutate(expression = as.numeric(expression))

# Remove rows with N/A
RNA_seq_tidy <- na.omit(RNA_seq_tidy)

# Convert column names to lower case (best practice)
colnames(RNA_seq_tidy) <- tolower(colnames(RNA_seq_tidy))

# Rename "sample" column to "Patient Barcode"
RNA_seq_tidy <- RNA_seq_tidy %>%
      rename("patient_barcode" = sample)

# Move 'patient_barcode' column to index 1
RNA_seq_tidy <- RNA_seq_tidy %>%
      select("patient_barcode", everything())

# Standardize format of patient barcodes from TCGA.A1.A0SD to TCGA-A1-A0SD (consistent across all datasets)
RNA_seq_tidy$patient_barcode <- gsub("\\.", "-", RNA_seq_tidy$patient_barcode)

colnames(RNA_seq_tidy)

# Change "gene name" to "gene_name"
RNA_seq_tidy <- RNA_seq_tidy %>%
      rename("gene_name" = 'attrib_name')

# Change "expression" to "expression_log2rsem"
RNA_seq_tidy <- RNA_seq_tidy %>%
      rename("expression_log2rsem" = 'expression')

colnames(RNA_seq_tidy)
nrow(RNA_seq_tidy) #3,233,933 rows?
head(RNA_seq_tidy)
View(RNA_seq_tidy)

# RNA-seq now in long/tidy format, i.e.,
# - Row = Patient
# - Columns = Gene expression variables (or Gene + Expression)

# Save Matrix as .csv
write.csv(RNA_seq_tidy, "RNA_seq_tidy.csv", row.names = FALSE)

# Load Matrix
RNA_seq_tidy <- read.csv("RNA_seq_tidy.csv")

----------------------------
# HISTOGRAM OF LOG2-TRANSFORMED GENE EXPRESSION

# Plot Distribution of Log2-Transformed Gene Expression Histogram
log2_histogram <- ggplot(RNA_seq_tidy, aes(x = expression)) +
                        geom_histogram(bins = 50, fill = "steelblue", color = "black") +
                        theme_minimal() +
                        labs(
                              title = "Distribution of Log2-Transformed Gene Expression",
                              x = "Log2 Expression (RSEM)",
                              y = "Frequency"
                        )

print(log2_histogram)

# Save Distribution of Log2-Transformed Gene Expression Histogram 
ggsave(
      filename = "log2_histogram.png",
      plot = log2_histogram,
      width = 8,
      height = 6,
      dpi = 300
)

----------------------------
IDENTIFICATION OF HIGHLY EXPRESSED GENES (NOT NECESSARILY CANCER RELATED)

# Identification of Breast Cancer Genes
b_cancer_genes <- (RNA_seq_tidy$gene_name)
# print(b_cancer_genes) #prints only column 1 of RNA Seq Data (i.e., the genes involved in BCancer)
length(b_cancer_genes) 

# Identification of unique genes
unique_genes <- unique(RNA_seq_tidy$gene_name)
length(unique_genes) # 6,149 UNIQUE genes
# print(unique_genes) # They're all unique

# Save Matrix as .csv
write.csv(unique_genes, "unique_genes.csv", row.names = FALSE)

# Load Matrix
unique_genes <- read.csv("unique_genes.csv")

----------------------------
# MEAN GENE EXPRESSION ANALYSIS AND GENE RANKING

# Calculate Mean Expression per gene
mean_gene_expression_dframe <- RNA_seq_data %>%
      rowwise() %>% 
      mutate(mean_expression = mean(c_across(-gene_name), na.rm = TRUE)) %>%
      ungroup() %>%
      select(gene_name, mean_expression) 

# TOP 20 Highly Expressed Breast Cancer Genes
top_20_exp_genes <- mean_gene_expression_dframe %>%
      arrange(desc(mean_expression)) %>% #sort gene expression means in decreasing order

      slice_head(n = 20)

View(top_20_exp_genes)
print(top_20_exp_genes)

# Save Matrix as .csv
write.csv(top_20_exp_genes, "top_20_exp_genes.csv", row.names = FALSE)

# Load Matrix
top_20_exp_genes <- read.csv("top_20_exp_genes.csv")

----------------------------
# VISUALIZATION & INTERPRETATION OF HIGHLY EXPRESSED GENES

# BARPLOT VISUALIZATION OF THE TOP 20 HIGHLY EXPRESSED GENES

# Create Barplot of the Top 20 Highly Expressed Genes
top20_exp_plot <- ggplot(top_20_exp_genes, aes(x = reorder(gene_name, mean_expression), y = mean_expression)) +
            geom_bar(stat = "identity", fill = "limegreen") +
            coord_flip() + #rearrange the coordinates so that we can easily read the gene names
            labs(
                  title = "Top 20 Highest Expressed Breast Cancer Genes",
                  x = "Genes",
                  y = "Mean Gene Expression (Log2 RSEM)"
            ) +
            theme_minimal() +
            theme(
                  axis.text.x = element_text(angle = 0, hjust = 0.5), #x-axis labels
                  axis.text.y = element_text(size = 10),              #y-axis labels (gene names)
                  axis.title.x = element_text(size = 12, face = "bold"),
                  axis.title.y = element_text(size = 12, face = "bold"),
                  plot.title = element_text(size = 14, face = "bold", hjust = 0.5) 
      )

print(top20_exp_plot)

# Note = These Top 20 highly expressed genes could be housekeeping- and other types of genes that do not have any prognostic significance.
# Next is the ID of survival-associated genes

# Save Top 20 Plot as .png
ggsave("Top_20_Highest_Expressed_Breast_Cancer_Genes.png", plot = top20_exp_plot, width = 6, height = 6, dpi = 300)

# Save as R Object
save(top20_exp_plot, file = "top20_exp_plot.RData")
load("top20_exp_plot.RData")

----------------------------
# ALTHOUGH I HAVE IDENTIFIED THE TOP 20 HIGHEST EXPRESSED GENES IN THIS BCANCER STUDY, 
# IT DOESN'T NECESSARILY MEAN THAT THESE TOP 20 GENES HAVE BIOLOGICAL OR PROGNOSTIC IMPORTANCE (I.E., THEY ARE NOT NECESSARILY SURVIVAL-ASSOCIATED GENES).
# THEY COULD BE HOUSEKEEPING GENES, IMMUNE FILTRATION GENES, OR STROMAL GENES THAT ARE 
# HIGHLY EXPRESSED ACROSS ALL TUMORS. 

# THE NEXT STEP AFTER THIS EXPLORATORY ANALYSIS STEP IS TO PREP. THE DATA FOR PCA & limma.
----------------------------

# DATA PREPARATION FOR DOWNSTREAM MODELING. 

lim_matrix <- as.matrix(RNA_seq_data[, -1])
rownames(lim_matrix) <- RNA_seq_data$gene_name

# Expression matrix
lim_matrix <- as.matrix(RNA_seq_data[, -1])
rownames(lim_matrix) <- RNA_seq_data$gene_name

pam50 <- factor(final_aggregated_patient_data$pam50) #Subset the PAM50 column from final_aggregated_patient_data and convert to factor (so that R knows we are dealing with categorical data, as opposed to continuous data).
names(pam50) <- final_aggregated_patient_data$patient_barcode #Assign PAM50 labels to respective patient barcodes (LumA, LumB, HER2, Basal)

# Remove NAs from PAM50 patients 
pam50 <- pam50[!is.na(pam50)]

# Match PAM50 samples with respective RNA-seq data via patient barcode
common_samples <- intersect(colnames(lim_matrix), names(pam50)) #Common samples = Patients that have both PAM50 labels and RNA_seq data
lim_matrix <- lim_matrix[, common_samples]
pam50 <- pam50[common_samples]
pam50 <- droplevels(pam50) #removes unused subtypes from the data. 

ncol(lim_matrix)  #445 - Not all patients have both PAM50 data and RNA_seq data. 
View(lim_matrix)

# Ensure Alignment
all(colnames(lim_matrix) == names(pam50))

# Design Matrix
design <- model.matrix(~0 + pam50)
colnames(design) <- levels(pam50)
rownames(design) <- names (pam50) #Add patient barcode to design matrix

View(design)
nrow(design) #445 - Not all patients have both PAM50 data and RNA_seq data.

# Remember for limma, ncol(lim_matrix) == nrow(design)

----------------------------
# PRINCIPAL COMPONENT ANALYSIS(PCA) ON RNA-SEQ DATA

gene_variance <- apply(lim_matrix, 1, var)

# Select the top 500 most variable genes
top_genes <- names(sort(gene_variance, decreasing = TRUE))[1:500]

# Transpose matrix so rows = samples and columns = genes
pca_input_top <- t(lim_matrix[top_genes, ])

pca_top <- prcomp(pca_input_top, scale. = TRUE)

# Convert output of PCA to a dataframe
pca_top_df <- as.data.frame(pca_top$x)

# Adds pam50 labels
pca_top_df$PAM50 <- pam50

# PCA Plot of Top 100 Variable Genes
pca_plot <- ggplot(pca_top_df, aes(PC1, PC2, color = PAM50)) +
                  geom_point(size = 2) +
                  theme_minimal() +
                  labs(title = "PCA of Top 500 Variable Genes")

print(pca_plot)

# Save PCA Plot of Top 100 Variable Genes
ggsave("PCA Plot.png", plot = pca_plot, width = 8, height = 6, dpi = 300)

----------------------------
# SCREE PLOT 

var_explained <- pca_top$sdev^2 / sum(pca_top$sdev^2) #explanation of the variance

# Scree dataframe
scree_df <- data.frame(
      PC = 1:10, #create dataframe of first 10 principal components.
      Variance = var_explained[1:10]
)

# Scree plot
scree_plot <- ggplot(scree_df, aes(x = PC, y = Variance)) +
      geom_line() +
      geom_point() +
      labs(
            title = "Scree Plot",
            x = "Principal Component",
            y = "Variance Explained"
      ) +
      theme_minimal()

print(scree_plot)

# Save Scree Plot as .png
ggsave("Scree Plot.png", plot = scree_plot, width = 8, height = 6, dpi = 300)

----------------------------
----------------------------
# DIFFERENTIAL GENE EXPRESSION ANALYSIS WITH LIMMA

# Afer ID of these significant genes, Cox regression (survival analysis) will be conducted
# on said subset of genes (i.e., only the biologically significant genes)

# Fit the linear model
fit <- lmFit(lim_matrix, design) #lmFit performs a linear regression on each of the 6149 genes to find differences in gene expression across PAM50 subtypes. 
fit <- eBayes(fit)

# Save design_matrix as .csv
write.csv(design, "design_matrix.csv", row.names = TRUE)

# Load Matrix
design <- read.csv("design_matrix.csv")

# Save lim_matrix as .csv
write.csv(lim_matrix, "lim_matrix.csv", row.names = TRUE)

# Load Matrix
lim_matrix <- read.csv("lim_matrix.csv")

----------------------------
# SUBTYPE-SPECIFIC DIFFERENTIAL EXPRESSION

# Gene Expression Estimates per Subtype
head(fit$coefficients)

# Create a ranked table of the genes that are most likely to be differentially expressed
diff_exp_results <- topTable(fit, number = Inf, adjust.method = "BH") #BH = Benjamini-Hochberg procedure = controls false discovery rate (i.e., the proportion of false positives among significant genes)

# Convert rownames to a column called gene names
diff_exp_results <- rownames_to_column(diff_exp_results, var = "gene_name")

head(diff_exp_results)
View(diff_exp_results)

# Differentially Expressed Genes (FDF < 0.05)
sum(diff_exp_results$adj.P.Val < 0.05) #6,047 #suggests that all but 102 genes are differentially expressed. 

# See below for calculation of genes differentially expressed across subtype contrasts (FDF < 0.05)
# 1,565-4,348 significantly different genes per subtype comparison (FDR <0.05). 

# Save subtype specific differential expression results as .csv
write.csv(diff_exp_results, "diff_exp_results.csv", row.names = TRUE)

# Load Matrix
diff_exp_results <- read.csv("diff_exp_results.csv")

----------------------------
----------------------------
# PAIR-WISE COMPARISONS OF GENE EXPRESSION.

      - Basal_vs_LumA
      - Basal_vs_LumB
      - Basal_vs_Her2
      - LumA_vs_LumB *
      - LumA_vs_Her2 
      - LumB_vs_Her2

# Pair-wise PAM50 Subtype Contrasts 
contrast.matrix <- makeContrasts(
      LumA_vs_Basal = LumA - Basal,
      LumB_vs_Basal = LumB - Basal,
      Her2_vs_Basal = Her2 - Basal,
      LumA_vs_LumB = LumA - LumB,
      LumA_vs_Her2 = LumA - Her2,
      LumB_vs_Her2 = LumB - Her2,
      levels = design
)

# contrast.matrix <- makeContrasts(
#       LumA_vs_Basal = pam50LumA - pam50Basal,
#       LumB_vs_Basal = pam50LumB - pam50Basal,
#       Her2_vs_Basal = pam50Her2 - pam50Basal,
#       LumA_vs_LumB = pam50LumA - pam50LumB,
#       LumA_vs_Her2 = pam50LumA - pam50Her2,
#       LumB_vs_Her2 = pam50LumB - pam50Her2,
#       levels = design
# )

fit2 <- contrasts.fit(fit, contrast.matrix) #Apply contrasts
fit2 <- eBayes(fit2) #Stabilizes gene variance estimates across all genes, thus improving statistical power = more reliable p-values.

# F-tests
diff_exp_pair_results <- topTableF(fit2, number = Inf) #Shows the total # of genes differentially expressed across all subtypes. topTableF performs ANOVA-style F-test across all PAM50 subtypes. 
sum(diff_exp_pair_results$adj.P.Val < 0.05) #5,100 of 6,149 genes total.

head(diff_exp_pair_results)
View(diff_exp_pair_results)

# Shows the total # of genes differentially expressed across subtype contrasts
LumA_vs_Basal_results <- topTable(fit2, coef = "LumA_vs_Basal", number = Inf)
LumB_vs_Basal_results <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf)
Her2_vs_Basal_results <- topTable(fit2, coef = "Her2_vs_Basal", number = Inf)
LumA_vs_LumB_results <- topTable(fit2, coef = "LumA_vs_LumB", number = Inf)
LumA_vs_Her2_results <- topTable(fit2, coef = "LumA_vs_Her2", number = Inf)
LumB_vs_Her2_results <- topTable(fit2, coef = "LumB_vs_Her2", number = Inf)

# Calculation of genes differentially expressed across subtype contrasts (FDF < 0.05)
sum(LumA_vs_Basal_results$adj.P.Val < 0.05) #4348 genes differentially expressed across LumA_vs_Basal
sum(LumB_vs_Basal_results$adj.P.Val < 0.05) #3740 genes differentially expressed across LumB_vs_Basal
sum(Her2_vs_Basal_results$adj.P.Val < 0.05) #2609 genes differentially expressed across Basal_vs_Her2
sum(LumA_vs_LumB_results$adj.P.Val < 0.05) #2687 genes differentially expressed across LumA_vs_LumB
sum(LumA_vs_Her2_results$adj.P.Val < 0.05) #2564 genes differentially expressed across LumA_vs_Her2
sum(LumB_vs_Her2_results$adj.P.Val < 0.05) #1565 genes differentially expressed across LumB_vs_Her2

# Save subtype-specific differential expression pair-wise results as .csv
write.csv(diff_exp_pair_results, "diff_exp_pair_results.csv", row.names = TRUE)

# Load Matrix
diff_exp_pair_results <- read.csv("diff_exp_pair_results.csv")

----------------------------
----------------------------
# PAIR-WISE CONTRAST MATRICES

# LumA v. Basal Contrast Matrix

lumA_basal <- topTable(fit2, coef = "LumA_vs_Basal", number = Inf)

# Add gene name to lumA_basal matrix
lumA_basal <- lumA_basal %>% rownames_to_column(var = "gene_name")

View(lumA_basal)

# Interpretation
# Positive logFC (e.g., logFC = 2), means the gene is expressed 4x more in LumA tumors than Basal tumors 
# Negative logFC (e.g., logFC = -2), means gene is express 4x more in Basal tumors than LumA tumors

# Save lumA_basal_matrix as .csv
write.csv(lumA_basal, "lumA_basal_matrix.csv", row.names = TRUE)

# Load Matrix
lumA_basal <- read.csv("lumA_basal_matrix.csv")

----------------------------
# LumA v. Her2 Contrast Matrix

lumA_Her2 <- topTable(fit2, coef = "LumA_vs_Her2", number = Inf)

# Add gene name to lumA_Her2 matrix
lumA_Her2 <- lumA_Her2 %>% rownames_to_column(var = "gene_name")

View(lumA_Her2)

# Interpretation
# Positive logFC = higher gene expression in LumA. 
# Negative logFC = higher gene expression in Her2.

# Save lumA_her2_matrix as .csv
write.csv(lumA_Her2, "lumA_her2_matrix.csv", row.names = TRUE)

# Load Matrix
lumA_Her2 <- read.csv("lumA_her2_matrix.csv")

----------------------------
# LumB v. Basal Contrast Matrix

lumB_basal <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf)

# Add gene name to lumB_basal matrix
lumB_basal <- lumB_basal %>% rownames_to_column(var = "gene_name")

View(lumB_basal)

# Interpretation
# Positive logFC = higher gene expression in LumB.
# Negative logFC = higher gene expression in Basal.

# Save lumA_her2_matrix as .csv
write.csv(lumB_basal, "lumB_basal_matrix.csv", row.names = TRUE)

# Load Matrix
lumB_basal <- read.csv("lumB_basal_matrix.csv")

----------------------------
# Her2 vs. Basal Contrast Matrix

Her2_basal <- topTable(fit2, coef = "Her2_vs_Basal", number = Inf)

# Add gene name to Her2_basal matrix
Her2_basal <- Her2_basal %>% rownames_to_column(var = "gene_name")

View(Her2_basal)

# Interpretation:
# Positive logFC = higher gene expression in Her2.
# Negative logFC = higher gene expression in Basal.

# Save Her2_Basal_matrix as .csv
write.csv(Her2_basal, "Her2_vs_Basal_matrix.csv", row.names = TRUE)

# Load Matrix
Her2_vs_Basal <- read.csv("Her2_vs_Basal_matrix.csv")

----------------------------
LumB v. Her2 Contrast Matrix

# LumB_basal contrast
lumB_Her2 <- topTable(fit2, coef = "LumB_vs_Her2", number = Inf)

# Add gene name to lumA_Her2 matrix
lumB_Her2 <- lumB_Her2 %>% rownames_to_column(var = "gene_name")

View(lumB_Her2)

# Interpretation
# positive logFC = higher gene expression in lumB.
# negative logFC = higher gene expression in Her2.

# Save lumA_her2_matrix as .csv
write.csv(lumB_Her2, "lumB_Her2.csv", row.names = TRUE)

# Load Matrix
lumB_Her2 <- read.csv("lumB_Her2_matrix.csv")

----------------------------
----------------------------
# DIFFERENTIAL EXPRESSION FILTERING & IDENTIFICATION OF THE BIOLOGICALLY & STATISTICALLY SIGNIFICANT GENES

# (i.e., significantly differentially expressed (DE) genes with False Discovery Rate (FDR) <0.05).
# FDR < 0.05 limits false positives (i.e., genes that appear significant by chance, but are not).
# This is done by filtering the DE genes by adjusted p-value. 

# GENES UPREGULATED IN LumA SUBTYPE.
lumA_sig_genes <- lumA_basal %>%
      filter(adj.P.Val < 0.05 & abs(logFC) > 1)

nrow(lumA_sig_genes) #654 significant genes
View(lumA_sig_genes)

# Upregulated genes in LumA subtype
lumA_upreg_genes <- lumA_sig_genes %>%
      filter(logFC > 1)

# 337 upregulated genes in LumA subtype
nrow(lumA_upreg_genes)

# Top 10 LumA upregulated genes
head(lumA_upreg_genes, 10)

print(lumA_upreg_genes)
View(lumA_upreg_genes)

# Save Upregulated LumA Genes as .csv
write.csv(lumA_upreg_genes, "lumA_upreg_genes.csv", row.names = TRUE)

# Load Matrix
lumA_upreg_genes <- read.csv("lumA_upreg_genes.csv")

----------------------------
# GENES UPREGULATED IN LumB SUBTYPE.

# Differential expression (DE) results for lumB vs. Basal subtypes
lumB_basal <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf) %>%
      rownames_to_column(var == "gene_name")

# Filtering for Significant DE genes
lumB_sig_genes <- lumB_basal %>%
      filter(adj.P.Val < 0.05 & abs(logFC) > 1)

lumB_upreg_genes <- lumB_sig_genes %>% filter(logFC >1)

# 248 Upregulated genes in LumB subtype
nrow(lumB_upreg_genes)

# Top 10 LumB Upregulated Genes
head(lumB_upreg_genes, 10)

print(lumB_upreg_genes)
View(lumB_upreg_genes)

# Save Upregulated LumB Genes as .csv
write.csv(lumB_upreg_genes, "lumB_upreg_genes.csv", row.names = TRUE)

# Load Natrix
lumB_upreg_genes <- read.csv("lumB_upreg_genes.csv")

----------------------------
# GENES UPREGULATED IN BASAL SUBTYPE.

basal_upreg_genes <- lumA_basal %>%
      filter(logFC < -1)

lumA_sig_genes <- lumA_basal %>%
      filter(adj.P.Val < 0.05 & abs(logFC) > 1)

basal_upreg_genes <- lumA_sig_genes %>%
      filter(logFC < -1)

# 317 Upregulated Genes in Basal Subtype
nrow(basal_upreg_genes)

# Top 10 Basal Upregulated Genes
head(basal_upreg_genes, 10)

# Print Basal Upregulated Genes
print(basal_upreg_genes)

# View Basal Upregulated Genes in Separate Window
View(basal_upreg_genes)

# Save Basal Upregulated Genes as .csv
write.csv(basal_upreg_genes, "basal_upreg_genes_matrix.csv", row.names = TRUE)

# Load Matrix
basal_upreg_genes <- read.csv("basal_upreg_genes_matrix.csv")

----------------------------
# GENES UPREGULATED IN Her2 SUBTYPE.

lumA_her2_sig_genes <- lumA_Her2 %>%
      filter(adj.P.Val < 0.05 & abs(logFC) >1)

her2_upreg_genes <- lumA_her2_sig_genes %>%
      filter(logFC < -1)

# 114 Upregulated Genes in Her2 Subtype
nrow(her2_upreg_genes)

# Top 10 Her2 Upregulated Genes
head(her2_upreg_genes, 10)

# Print Top 10 Her2 Upregulated Genes
print(her2_upreg_genes)

# View Top 10 Her2 Upregulated Genes in Separate Window
View(her2_upreg_genes)

# Save Upregulated Basal Genes as .csv
write.csv(her2_upreg_genes, "her2_upreg_genes.csv", row.names = TRUE)

# Load Matrix
her2_upreg_genes <- read.csv("her2_upreg_genes.csv")

----------------------------
----------------------------
# VOLCANO PLOTS TO VISUALIZE GENES THAT ARE UPREGULATED AND DOWNREGULATED ACROSS SUBTYPES.

# LumA v. Basal Volcano Plot
top_genes <- lumA_basal %>% #Subsets the most significant genes to label
      arrange(adj.P.Val) %>% #Sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) #First 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

# Create LumA v. Basal Volcano Plot
lumA_vs_basal_volcano_plot <- ggplot(lumA_basal, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      annotate("text", x = Inf, y = -log10(0.05), label = "p-value = 0.05", hjust = 1.1, vjust = -1.0, colors = "red", size = 4) +            
      theme_minimal() +
      labs(
            title = "LumA v. Basal Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()

# Print LumA v. Basal Volcano Plot
print(lumA_vs_basal_volcano_plot)

# Save LumA v. Basal Volcano Plot as .png 
ggsave("lumA_vs_basal_volcano_plot.png", plot = lumA_vs_basal_volcano_plot, width = 8, height = 6, dpi = 300)

----------------------------
# LumA v. Her2 Volcano Plot

# LumA_Her2 contrast
lumA_Her2 <- topTable(fit2, coef = "LumA_vs_Her2", number = Inf)

# Add gene name to lumA_Her2 Matrix
lumA_Her2 <- lumA_Her2 %>% rownames_to_column(var = "gene_name")

top_genes_lumA_her2 <- lumA_Her2 %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumA_vs_her2_volcano_plot <- ggplot(lumA_Her2, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_lumA_her2, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      annotate("text", x = Inf, y = -log10(0.05), label = "p-value = 0.05", hjust = 1.1, vjust = -1.0, colors = "red", size = 4) +      
      theme_minimal() +
      labs(
            title = "LumA v. Her2 Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()

# Print LumA v. Her2 Volcano Plot
print(lumA_vs_her2_volcano_plot)

# Save LumA v. Her2 Volcano Plot as .png
ggsave("lumA_vs_her2_volcano_plot.png", plot = lumA_vs_her2_volcano_plot, width = 8, height = 6, dpi = 300)

----------------------------
# LumB v. Basal Volcano Plot

# LumB_basal Contrast
lumB_basal <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf)

# Add gene name to lumB_basal matrix
lumB_basal <- lumB_basal %>% rownames_to_column(var = "gene_name")

top_genes_lumB_basal <- lumB_basal %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumB_vs_basal_volcano_plot <- ggplot(lumB_basal, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_lumB_basal, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      annotate("text", x = Inf, y = -log10(0.05), label = "p-value = 0.05", hjust = 1.1, vjust = -1.0, colors = "red", size = 4) +
      theme_minimal() +
      labs(
            title = "LumB v. Basal Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()

# Print LumB v. Basal Volcano Plot
print(lumB_vs_basal_volcano_plot)

# Save LumB v. Basal Volcano Plot as .png
ggsave("lumB_vs_basal_volcano_plot.png", plot = lumB_vs_basal_volcano_plot, width = 8, height = 6, dpi = 300)

----------------------------
# Her2 v. Basal Volcano Plot

# LumB_basal Contrast
Her2_basal <- topTable(fit2, coef = "Her2_vs_Basal", number = Inf)

# Add gene name to lumA_Her2 matrix
Her2_basal <- Her2_basal %>% rownames_to_column(var = "gene_name")

top_genes_Her2_basal <- Her2_basal %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

Her2_vs_basal_volcano_plot <- ggplot(Her2_basal, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_Her2_basal, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      annotate("text", x = Inf, y = -log10(0.05), label = "p-value = 0.05", hjust = 1.1, vjust = -1.0, colors = "red", size = 4) +
      theme_minimal() +
      labs(
            title = "Her2 v. Basal Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz() #Opens plot in new window

# Print Her2 v. Basal Volcano Plot
print(Her2_vs_basal_volcano_plot)

# Save Her2 v. Basal Volcano Plot as .png
ggsave("Her2_vs_basal_volcano_plot.png", plot = Her2_vs_basal_volcano_plot, width = 8, height = 6, dpi = 300)

----------------------------
# LumB v. Her2 Volcano Plot

# LumB_Her2 contrast
lumB_Her2 <- topTable(fit2, coef = "LumB_vs_Her2", number = Inf)

# Add gene name to lumA_Her2 matrix
lumB_Her2 <- lumB_Her2 %>% rownames_to_column(var = "gene_name")

top_genes_lumB_Her2 <- lumB_Her2 %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumB_vs_Her2_volcano_plot <- ggplot(lumB_Her2, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_lumB_Her2, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      annotate("text", x = Inf, y = -log10(0.05), label = "p-value = 0.05", hjust = 1.1, vjust = -1.0, colors = "red", size = 4) +      
      theme_minimal() +
      labs(
            title = "LumB v. Her2 Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()

# Print LumB v. Her2 Volcano Plot
print(lumB_vs_Her2_volcano_plot)

# Save LumB v. Her2 Volcano Plot
ggsave("lumB_vs_Her2_volcano_plot.png", plot = lumB_vs_Her2_volcano_plot, width = 8, height = 6, dpi = 300)

----------------------------
----------------------------
# HEATMAP OF GENE EXPRESSION PATTERNS/HEATMAP OF TOP VARIABLE GENES ACROSS PAM50 SUBTYPES.

# Calculate the variance for each gene
gene_variances <- apply(lim_matrix, 1, var, na.rm = TRUE)

# Select the top 100 most variable genes
top_var_genes <- names(sort(gene_variances, decreasing = TRUE))[1:100]

# Subset the top 100 variable genes
heatmap_data <- lim_matrix[top_var_genes, ]

# Scales the data by standardizing each gene (row) - enhances visualization
heatmap_scaled <- t(scale(t(heatmap_data)))

# Add pam50 annotations
annotation_col <- data.frame(PAM50 = pam50)
rownames(annotation_col) <- colnames(heatmap_scaled)

# Plot & Save the Heatmap
png("Heatmap_of_Top_100_Variable_Genes.png", width = 800, height = 600, res = 150)
heatmap_plot <- pheatmap(
                        heatmap_scaled,
                        annotation_col = annotation_col,
                        show_rownames = FALSE,
                        show_colnames = FALSE,
                        clustering_method = "complete",
                        scale = "none",
                        main = "Heatmap of Top 100 Variable Genes Across PAM50 Subtypes"
                  )
dev.off()

print(heatmap_plot)

----------------------------
----------------------------
# MERGE GENE EXPRESSION WITH SURVIVAL DATA

# Rows = patient
# Columns = gene expression + survival data

# Merge RNA-seq data and Survival Data
gene_expr <- as.data.frame(t(RNA_seq_data))
colnames(gene_expr) <- RNA_seq_data$gene_name #preseve gene names 
gene_expr$patient_barcode <- rownames(gene_expr)

survival_dataset <- dplyr::left_join(
      final_aggregated_patient_data,
      gene_expr,
      by = "patient_barcode"
)

# View(as.data.frame(survival_dataset))
View(survival_dataset)

# "overall_survival" %in% colnames(survival_dataset) #true
# "status" %in% colnames(survival_dataset) #true
# "demographic.vital_status" %in% colnames(survival_dataset) #true
# "demographic.days_to_death" %in% colnames(survival_dataset) #true

# Save Upregulated survival_dataset as .csv
write.csv(survival_dataset, "survival_dataset.csv", row.names = TRUE)

# Load Matrix
survival_dataset <- read.csv("survival_dataset.csv")

----------------------------
# CREATE COLUMNS FOR SURVIVAL ANALYSIS

#      - Censoring Status = values in this column are either:
#            - 1 (i.e., Event Happened (i.e., Death)) #151
#            - 0 (i.e., Censored (i.e., Alive)) #843
#      - Time-To-Event = from the start time (i.e., time of diagnosis), to the end time (i.e., time when event (death) occured or censoring (i.e., last follow-up))

survival_dataset <- survival_dataset %>%
      mutate(
            follow_up_days = as.numeric(diagnoses.days_to_last_follow_up), #for patients who are still alive/censored
            death_days = as.numeric(demographic.days_to_death), #for patients who have died
            time_to_event = ifelse(!is.na(death_days), death_days, follow_up_days),
            event_status = as.numeric(status)
      ) %>%
      filter(!is.na(time_to_event)) #filters patients with no follow-up time or no death-time

# table(survival_dataset$event_status)
table(survival_dataset$event_status, useNA = "always") # 0(alive) = 843, 1(deaths) = 151, NAs = 42
summary(survival_dataset$time_to_event)

# Remove 42 NA's from event_status
survival_dataset <- survival_dataset %>%
      filter(!is.na(event_status))

View(survival_dataset)

----------------------------
# SURVIVAL OBJECT

surv_obj <- Surv(
      time = survival_dataset$time_to_event,
      event = survival_dataset$event_status
)

View(surv_obj) 
# 0(alive) = 843, 1(deaths) = 151, NAs = 42

----------------------------
----------------------------
# BUILD COX MODEL

# Evaluate multiple gene signatures. 
# Build kaplan on gene signature with highest prognostic ability. 

----------------------------
# Cox Model for BCL2 only

# Extract the BCL2 column from the survival dataset
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, BCL2) %>%
      na.omit()

cox_data$BCL2 <- as.numeric(trimws(cox_data$BCL2)) #ensures that the BCL2 column is numeric.

# Run Cox Model on BCL2 only
cox_model <- coxph(
      Surv(time_to_event, event_status) ~ BCL2,
      data = cox_data
)

summary(cox_model)

# Calculate Hazard Ratio
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

----------------------------
# Cox Model for ARBITRARY GENE SIGNATURE using Top DEGs
genes_of_interest <- c("BCL2", "FOXA1", "UGT8", "ROPN1", "HORMAD1", "CA12", "SERPINA3", "GREB1", "MAPT", "TBC1D9", "SLC44A4", "LRFN2", "GSDML", "C17orf37", "PNMT", "RRM2", "GRB7")

# Subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(genes_of_interest)) %>%
      na.omit() #remove NA's

# Make the gene columns numeric
cox_data[genes_of_interest] <- lapply(cox_data[genes_of_interest], function(x) as.numeric(trimws(x)))

# Run multivariate cox model on said 17 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(genes_of_interest, collapse = " + "))),
      data = cox_data
)

summary(cox_model)

# Calculate Hazard Ratio
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

----------------------------
# BASAL GENE SIGNATURE
# Cox model on 6 genes upreg in Basal subtype (from limma results)

# Genes Upregulated in the Basal subtype
up_genes_basal <- c("ROPN1", "UGT8", "PPP1R14C", "HORMAD1", "KRT16", "ROPN1B")

# Subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(up_genes_basal)) %>%
      na.omit() #remove NA's

# Make the gene columns numeric
cox_data[up_genes_basal] <- lapply(cox_data[up_genes_basal], function(x) as.numeric(trimws(x)))

# Run multivariate cox model on said 6 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(up_genes_basal, collapse = " + "))),
      data = cox_data
)

summary(cox_model)

# Calculate Hazard Ratio
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

----------------------------
# HER2 GENE SIGNATURE
# Cox model on  genes upregulated in Her2 subtype

# Genes Upregulated in Her2 subtype
up_genes_her2 <- c("GRB7", "PNMT", "RRM2", "GSDML", "C17orf37", "CLCA2", "ATP13A5", "AKR1B10", "MLPH", "GPR160", "LRFN2")

# Subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(up_genes_her2)) %>%
      na.omit() #remove NA's

# Make the gene columns numeric
cox_data[up_genes_her2] <- lapply(cox_data[up_genes_her2], function(x) as.numeric(trimws(x)))

# Run multivariate cox model on said 11 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(up_genes_her2, collapse = " + "))),
      data = cox_data
)

summary(cox_model)
summary(cox_model)$coefficients

----------------------------
# LUMINAL GENE SIGNATURE
# Cox model on  genes upregulated in luminal subtypes (both lumA & lumB) (from limma results)

# Genes Upregulated in Luminal subtypes
up_genes_luminal <- c("BCL2", "SERPINA3", "GREB1", "MAPT", "FOXA1", "CA12", "TBC1D9", "MLPH", "SLC44A4", "PARD6B", "PPM1J", "LRFN2")

# Subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(up_genes_luminal)) %>%
      na.omit() #removes NA's

# Make the gene columns numeric
cox_data[up_genes_luminal] <- lapply(cox_data[up_genes_luminal], function(x) as.numeric(trimws(x)))

# Run multivariate cox model on said 12 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(up_genes_luminal, collapse = " + "))),
      data = cox_data
)

summary(cox_model)

# Calculate Hazard Ratio
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

----------------------------
----------------------------
# CALCULATION OF RISK (PROGNOSTIC) SCORE for LUMINAL GENE SIGNATURE

# A risk score is calculated for each patient.
# Risk scores are used to separate patients into HIGH RISK v. LOW RISK GROUPS.
# It is this prognostic score that turns the model to into a predictive tool. 

# Add Risk Scores
cox_data$risk_score <- predict(cox_model, type = "lp")

head(cox_data$risk_score)

positive_risk <- cox_data[cox_data$risk_score > 0, ]
nrow(positive_risk) #Number of patients with positive risk scores = 223

negative_risk <- cox_data[cox_data$risk_score < 0, ]
nrow(negative_risk) #Number of patients with negative risk scores = 245

----------------------------
# STRATIFY PATIENTS into "High Risk" v. "Low Risk" groups" using cox_data dataset

# Stratify Patients by Median Risk Score
median_score <- median(cox_data$risk_score)

cox_data <- cox_data %>%
      mutate(
            risk_group = ifelse(risk_score > median_score, "High Risk", "Low Risk")
      )

table(cox_data$risk_group)

----------------------------
# Interpretation:
#      - High risk = worse survival
#      - Low risk = better survival

----------------------------
----------------------------
# KAPLAN-MEIER SURVIVAL PLOT - Visualization of the comparison between "High Risk" v. "Low Risk" groups.

# Plotted using Luminal Multi-gene Signature

# Creation of Survival Object
surv_obj <- Surv(time = cox_data$time_to_event, event = cox_data$event_status)

# Fit kaplan-meier (high risk vs. low risk)
km_fit <- survfit(surv_obj ~ risk_group, data = cox_data)
# km_fit <- survfit(Surv(time_to_event, event_status) ~ risk_group, data = cox_data)

# Create Kaplan-Meier Plot
kaplan <- ggsurvplot(
      km_fit,
      data = cox_data,
      risk.table = TRUE, 
      pval = TRUE,
      conf.int = TRUE,
      palette = c("red", "green"), #distinguish low risk (green) vs. high risk (red) by colour
      xlab = "Time (days)",
      ylab = "Overall Survival Probability",
      title = "Kaplan-Meier Survival Curves by Risk Group",
      legend.title = "Risk Group",
      legend.labs = c("High Risk", "Low Risk") #label high and low risk groups
)

print(kaplan)

# Save Kaplan-Meier Plot
ggsave(
      filename = "KM_survival_plot.png",
      plot = kaplan$plot,
      width = 8,
      height = 6,
      dpi = 300
)

----------------------------
# ESTIMATE X-YEAR SURVIVAL (i.e., the probability of surviving after x number of years for patients in the high expression v. low expression groups)
summary(km_fit, times = c(365, 1095, 1825)) #time in days

----------------------------
# ESTIMATE MEDIAN SURVIVAL TIME (i.e., the estimated time when survival = 0.5)
summary(km_fit)

----------------------------
# UNIVARIATE COX REGRESSION ON RISK SCORE
# Evaluates if multi-gene signature (i.e., risk score) actually predicts survival - and quantifies.

coxph(Surv ~ risk_score)

cox_uni <- coxph(Surv(time_to_event, event_status) ~ risk_score,
      data = cox_data)

summary(cox_uni)

# Calculate Hazard Ratio
exp(coef(cox_uni)) #HR = 2.718282 




